---
title: "LLM analysis of a biosecurity corpus"
editor: source
date: today
bibliography: 
  - AI-biosecurity.bib
csl: apa-tr.csl

execute:
  echo: true
  warning: false
  message: false
  freeze: auto

format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    output-file: "index.html"
    fig-cap-location: margin
    tbl-cap-location: margin
    fig-width: 5.5

author:
  - name: Christopher E Buddenhagen
    affiliation: Bioeconomy Science Institute, Ruakura, Hamilton, New Zealand
---

```{r setup}

required_packages <- c(
  "tidyverse","readxl","janitor","httr","jsonlite",
  "progress","tictoc","openai","gt","knitr"
)

installed <- rownames(installed.packages())

for (pkg in required_packages) {
  if (!pkg %in% installed) install.packages(pkg)
}

invisible(lapply(required_packages, library, character.only = TRUE))
```

# Introduction


::: {.callout-important}
### Download Workshop Materials

All files needed for the Biosecurity Symposium workshop (University of Hawaii, 14-19 June 2026) are available here:

👉 https://github.com/cbud/biosecurity-symposium/archive/refs/heads/main.zip

Download and unzip the file, then open the project in RStudio.
:::

### Required files

This workshop uses:

- `corpus/` (text data)
- `functions/` (helper functions)
- `prompts/` (prompt templates)
- `AI biosecurity workshop generalized.R` (main script)

All are included in the download. You can generate your own API key (not free) at: \
<https://openai.com/api>. One can be provided during the workshop.


## LLM‑based extraction of biosecurity information

This document presents a generalised LLM pipeline for extracting structured information from bibliographic datasets. It supports scalable literature review and risk assessment workflows and was applied in rapid hazard ranking of invasive plant species [@buddenhagenValidatingRapidAlgorithmic2025].

The model uses three variables:

- scientific publication counts  
- climate suitability  
- species distribution  

to rank potential invasive taxa.

@buddenhagenValidatingRapidAlgorithmic2025 introduced a rapid, algorithmic weed hazard ranking model (CPG) that uses widely available data. It leverages two well-supported predictors of weediness: weediness elsewhere and climate suitability [@caleyApplicationEvaluationClassification2006; @hulmeWeedRiskAssessment2012; @weberAnalysisAssessmentOutcomes2009; @randallCanPlantsCultural2016].

Evidence shows that an organism’s invasion history (i.e., weediness elsewhere) generally outperforms biological traits (or attributes) as a predictor of risk. This is partly because selecting consistent trait-based variables is difficult across organism groups, and such data are often incomplete or unavailable.

The model estimates hazard by combining:

1. climatic suitability (C) in the receiving region  
2. the number of weed-related scientific publications (P), derived using LLMs as a proxy for impact and concern  
3. global occurrence counts (G), reflecting ecological generalism and spread potential  

High-scoring species can then be prioritised for more detailed risk assessments to evaluate potential harms, affected stakeholders, management feasibility, costs, and benefits.

### Steps

- Compile a dataframe with the corpus (text for querying with AI), with one row per source  
- Develop a prompt for use with an OpenAI model (e.g. GPT‑4o‑mini)  
- Run the prompts across the dataframe  
- Standardise names to the GBIF backbone  

## Background on Large Language Models (LLMs)

Many modern large language models (LLMs) can be accessed via APIs that use a *pay-as-you-go* pricing model. Costs are typically based on the number of **tokens** processed.

A *token* is a unit of text used internally by LLMs. It is not the same as a word; instead, it often corresponds to word pieces (for example, parts of words or short sequences of characters). This tokenization allows models to efficiently process and generate text.

When using an API, you are generally charged for: - **Input tokens** (text sent to the model), and\
- **Output tokens** (text generated by the model)

Pricing varies by model and provider. For example, OpenAI publishes current pricing here:\
<https://developers.openai.com/api/docs/pricing>

In this project, we use the `gpt-4o-mini` model due to its balance of cost, speed, and capability. It provides relatively concise outputs and seems to respond well to the constraints in the prompts. Other models ( e.g., newer releases like gpt-5 or higher) would require changes to the code and prompts, as APIs and model interfaces can evolve over time. I tried, it didn't work.

To access an LLM API, you also need an **API key**. This is a private credential that authenticates your requests and links usage to your account. It should be stored securely (e.g., in environment variables) and never committed to version control.

Similar API-based access patterns are available across multiple providers, including OpenAI, Anthropic, Google, and others.

# Running an example

## Example data set

This pipeline used bibliometric data downloaded from the Scopus GUI, the query structure was: TITLE-ABS-KEY ( weed\* OR invas\* OR invad\* OR "alien plant*" OR "non-native plant*" OR "exotic plant*" OR "introduced plant*" OR naturalised OR naturalized ) AND TITLE-ABS-KEY ( plant\* OR vegetation OR species ) AND NOT TITLE-ABS-KEY ( insect\* OR animal\* OR bacter\* OR virus\* ). The full scopus had 156,000 abstracts - these had to be downloaded 20,000 at a time. Processing this amount of data cost about \$40 USD and took three days. For the workshop I select a subset of 100 random articles with invasive plants or weeds mentioned and 100 that were determined by AI to be off-topic.

The species names (Latin binomials) in the abstract have been manually reviewed and are entered one species per row, with repeated bibliometric data if more than one species is mentioned in the abstract. To use the typical R language the data is long with respect to unique manually identified species.

```{r, warning=FALSE, echo=FALSE}

corpus<-read_xlsx("corpus/scopus_sample_data_set.xlsx") %>% distinct() %>% clean_names()

#You must have title_abstract field in your dataframe for this pipeline - in this case you can include keyword fields.

#dput(names(corpus))




```

The corpus contains `r length(unique(corpus$eid))` records and
`r length(unique(corpus$manual_calls_weeds))` manually identified species.


## Prompt set up

This AI pipeline uses a chat completion approach with an LLM to populate columns in a dataframe through structured prompts. Each prompt consists of two components: **system text** (instructions that define behaviour and constraints) and **user text** (the specific task applied to each record).

Prompts can be written manually or generated iteratively by asking an LLM to refine or expand an example prompt to cover multiple topics or variables. In this workflow, the initial prompt is designed to classify whether an article is relevant (“on topic”). Subsequent extraction prompts are only applied to records that pass this filter, reducing both runtime and token usage. This filtering step is optional but recommended for efficiency.

The prompt configuration is stored as a `topic_settings` object in the code, with the corresponding prompt defined in:
 "prompts/prompt_1.R".
 
 When setting up prompts, you should explicitly define:
- the task (e.g. classification, entity extraction, counting)
- the expected output format (e.g. “yes/no”, delimited lists, or structured fields)
- any constraints or definitions (e.g. what qualifies as “weed-related”)
- how to handle uncertainty (e.g. return “no” or “unknown”)

Clear, consistent instructions improve reproducibility and downstream parsing.


## Run the LLM‑based data recovery

To run the pipeline, configure the user settings that control data input/output, model behaviour, and processing workflow.

Key elements to specify:

- **API key**: stored in a local text file and referenced via `api_key_file`  
- **Input data**: ensure the dataframe includes a column named `title_abstract` containing the text to analyse  
- **Output directory**: where results will be written  
- **Model selection**: e.g. `gpt-4o-mini`, balancing cost and performance  
- **Chunk size**: splits processing into batches to improve stability and enable restart from partial runs  
- **Topic filter settings**: column name and pattern defining “on topic” records  
- **Entity extraction settings**: column name and delimiter for parsed outputs  (in this case we mean species - since it is tied to the taxonomy)
- **Taxonomic standardisation**: e.g. GBIF kingdom for downstream name matching  

Chunking is especially important for large datasets. For example, with ~1000 articles per chunk, each topic may take approximately 9–15 minutes. If a run fails, chunking allows you to resume without reprocessing completed records (with some changes to the script).

```{r AI extraction}
## AI biosecurity workshop generalized

# ===========================================================
# LLM ANALYSIS PIPELINE — GENERALISED SCRIPT
# ===========================================================

# -------------------------------
# USER SETTINGS (EDIT THIS BLOCK)
# -------------------------------

settings <- list(
  
  # INPUT DATA
  input_file = "corpus/scopus_sample_data_set.xlsx",
  
  # OUTPUT
  out_dir = "./outputs/example_ai_run",
  
  # MODEL + PROCESSING
  model = "gpt-4o-mini",
  chunk_size = 100,
  
  # TOPIC FILTER COLUMN (from topic_settings$output_column)
  on_topic_column = "on_topic_parsed",   # <-- changeable
  
  # STRING THAT DEFINES "YES"
  on_topic_pattern = "^yes",
  
  # ENTITY COLUMN (parsed names from AI)
  entity_column = "weed_name_parsed",            # <-- changeable
  entity_delim = ";",
  
  # GBIF SETTINGS
  gbif_kingdom = "Plantae",                      # change if needed
  
  # API KEY FILE
  api_key_file = "Biosecurity workshop api/Biosecurity_symposium_api.txt"
)

# -------------------------------
# LOAD DATA
# -------------------------------

corpus <- read_xlsx(settings$input_file) %>%
  distinct() %>%
  clean_names() # < may need to read in with another function if this is in a csv format Scopus uses for outputs

# Build text field
corpus <- corpus %>%
  rowwise() %>%
  mutate(
    title_abstract = stringr::str_squish(
      paste(
        na.omit(c(title, abstract, author_keywords, index_keywords)),
        collapse = " "
      )
    )
  ) %>%
  ungroup()

data_pubs_all <- corpus

data_pubs <- data_pubs_all %>%
  distinct(eid, .keep_all = TRUE) %>%
  mutate(row_id = row_number())

# ===========================================================
# API KEY
# ===========================================================

if (file.exists(settings$api_key_file)) {
  Sys.setenv(OPENAI_API_KEY = readLines(settings$api_key_file, warn = FALSE))
} else {
  stop("API key file not found.")
}

# ===========================================================
# LOAD PROMPTS + FUNCTIONS
# ===========================================================

source("prompts/prompt_1.R")
source("functions/f_ai_text_analysis.R")
source("functions/f_get_gbif_backbone_ids_from_df.R")

# -------------------------------
# BUILD title_abstract FIELD
# -------------------------------
corpus <- corpus %>%
  rowwise() %>%
  mutate(
    title_abstract = stringr::str_squish(
      paste(na.omit(c(title, abstract, author_keywords, index_keywords)),
            collapse = " ")
    )
  ) %>%
  ungroup()

data_pubs_all <- corpus

data_pubs <- data_pubs_all %>%
  distinct(eid, .keep_all = TRUE) %>%
  mutate(row_id = row_number())

# -------------------------------
# CREATE OUTPUT DIR
# -------------------------------
if (!dir.exists(settings$out_dir)) dir.create(settings$out_dir, recursive = TRUE)

# ===========================================================
# 1. RUN TOPIC FILTER (ROW 1 OF topic_settings)
# ===========================================================
tic()

topic_category <- topic_settings$topic_category[1]
output_column  <- topic_settings$output_column[1]
system_text    <- topic_settings$system_text[1]
user_text      <- topic_settings$user_text[1]

topic_dir <- file.path(settings$out_dir, topic_category)

# ✅ CLEAN DIRECTORY (prevents duplicate chunk stacking)
if (dir.exists(topic_dir)) unlink(topic_dir, recursive = TRUE)
dir.create(topic_dir, recursive = TRUE)

chunks <- split(data_pubs,
                ceiling(seq_len(nrow(data_pubs)) / settings$chunk_size))

for (i in seq_along(chunks)) {
  
  message("Processing ", topic_category, " | Chunk ", i)
  
  result_vector <- f_ai_text_analysis(
    chunks[[i]]$title_abstract,
    settings$model,
    system_text,
    user_text
  )
  
  chunks[[i]][[output_column]] <- result_vector
  
  saveRDS(
    chunks[[i]],
    file = file.path(topic_dir, paste0(topic_category, "_chunk_", i, ".rds"))
  )
  
  # ✅ MEMORY CLEANING
  rm(result_vector)
  gc()
}

# ✅ FREE CHUNKS FROM MEMORY
rm(chunks)
gc()

# ✅ LOAD ONLY chunk files
files <- list.files(topic_dir, full.names = TRUE, pattern = "\\.rds$")

on_topic_final <- purrr::map_dfr(files, readRDS) %>%
  distinct(eid, .keep_all = TRUE)   # ✅ safeguard against duplication

# ✅ CLEAN TEMP OBJECTS
rm(files)
gc()

saveRDS(
  on_topic_final,
  file = file.path(topic_dir, paste0(topic_category, "_final_results.rds"))
)

toc()

# -------------------------------
# FILTER ON-TOPIC (GENERALIZED)
# -------------------------------
on_topic_pubs <- on_topic_final %>%
  filter(stringr::str_detect(
    tolower(.data[[settings$on_topic_column]]),
    settings$on_topic_pattern
  ))

# ✅ CLEAN UNUSED OBJECT
rm(on_topic_final)
gc()

# ===========================================================
# 2. RUN REMAINING AI TASKS
# ===========================================================
tic()

for (j in 2:nrow(topic_settings)) {
  
  topic_category <- topic_settings$topic_category[j]
  output_column  <- topic_settings$output_column[j]
  system_text    <- topic_settings$system_text[j]
  user_text      <- topic_settings$user_text[j]
  
  topic_dir <- file.path(settings$out_dir, topic_category)
  
  # ✅ CLEAN DIRECTORY BEFORE RUN
  if (dir.exists(topic_dir)) unlink(topic_dir, recursive = TRUE)
  dir.create(topic_dir, recursive = TRUE)
  
  chunks <- split(on_topic_pubs,
                  ceiling(seq_len(nrow(on_topic_pubs)) / settings$chunk_size))
  
  for (i in seq_along(chunks)) {
    
    message("Processing ", topic_category, " | Chunk ", i)
    
    result_vector <- f_ai_text_analysis(
      chunks[[i]]$title_abstract,
      settings$model,
      system_text,
      user_text
    )
    
    chunks[[i]][[output_column]] <- result_vector
    
    saveRDS(
      chunks[[i]],
      file = file.path(topic_dir, paste0(topic_category, "_chunk_", i, ".rds"))
    )
    
    # ✅ MEMORY CLEANING (critical)
    rm(result_vector)
    gc()
  }
  
  # ✅ REMOVE CHUNKS FROM MEMORY
  rm(chunks)
  gc()
  
  files <- list.files(topic_dir, full.names = TRUE, pattern = "\\.rds$")
  
  final_df <- purrr::map_dfr(files, readRDS) %>%
    distinct(eid, .keep_all = TRUE)  # ✅ safeguard
  
  # ✅ CLEAN FILE LIST
  rm(files)
  gc()
  
  saveRDS(
    final_df,
    file = file.path(topic_dir, paste0(topic_category, "_final_results.rds"))
  )
  
  # ✅ FREE MEMORY BEFORE NEXT LOOP
  rm(final_df)
  gc()
}

toc()

```

After you run the ai recovery the data is stored for each chunk in separate folders for each topic and needs to be combined into a master file that stores the topic outputs described in the prompt and the original data. This code should work no matter how many prompt topics you are running. The script here takes about 5 minutes. An example script is also here: "AI biosecurity workshop generalized.R".

```{r combine ai chunks}
# ===========================================================
# MERGE AI RESULTS 
# ===========================================================

merge_ai_results <- function(topic_settings, out_dir, join_col = "eid") {
  
  results_list <- purrr::pmap(
    topic_settings,
    function(topic_category, output_column, system_text, user_text) {
      
      results_file <- file.path(
        out_dir,
        topic_category,
        paste0(topic_category, "_final_results.rds")
      )
      
      if (!file.exists(results_file)) return(NULL)
      
      readRDS(results_file) %>%
        distinct(.data[[join_col]], .keep_all = TRUE) %>%
        select(all_of(c(join_col, output_column)))
    }
  )
  
  purrr::compact(results_list)
}

ai_results <- merge_ai_results(topic_settings, settings$out_dir)

data_pubs_final <- purrr::reduce(
  ai_results,
  dplyr::left_join,
  by = "eid",
  .init = data_pubs_all
) %>%
  distinct()

saveRDS(data_pubs_final,
        file = file.path(settings$out_dir, "data_pubs_final.rds"))
```

## Recover GBIF standard taxonomy

In our pipelines we standardize all Latin binomials to the [GBIF backbone taxonomy](https://www.gbif.org/dataset/d7dddbf4-2cf0-4f39-9b2a-bb099caae36c#description). This involves making the data long (multiple taxa are separated by semicolons in each row) and then extracting the names from the GBIF API. If a species isn't resolved by the backbone extraction process it is stored in the output directory you specified: `r settings$out_dir`.

```{r taxonomy standardisation}


# ===========================================================
# MERGE AI RESULTS AND GET GBIF NAMES
# ===========================================================

merge_ai_results <- function(topic_settings, out_dir, join_col = "eid") {
  
  results_list <- purrr::pmap(
    topic_settings,
    function(topic_category, output_column, system_text, user_text) {
      
      results_file <- file.path(
        out_dir,
        topic_category,
        paste0(topic_category, "_final_results.rds")
      )
      
      if (!file.exists(results_file)) return(NULL)
      
      readRDS(results_file) %>%
        distinct(.data[[join_col]], .keep_all = TRUE) %>%
        select(all_of(c(join_col, output_column)))
    }
  )
  
  purrr::compact(results_list)
}

ai_results <- merge_ai_results(topic_settings, settings$out_dir)

data_pubs_final <- purrr::reduce(
  ai_results,
  dplyr::left_join,
  by = "eid",
  .init = data_pubs_all
) %>%
  distinct()

saveRDS(data_pubs_final,
        file = file.path(settings$out_dir, "data_pubs_final.rds"))

# ===========================================================
# LONG FORMAT + GBIF (FULLY GENERALISED)
# ===========================================================

data_pubs_long <- data_pubs_final %>%
  tidyr::separate_longer_delim(
    .data[[settings$entity_column]],
    delim = settings$entity_delim
  ) %>%
  mutate(
    entity_clean = stringr::str_squish(.data[[settings$entity_column]])
  ) %>%
  filter(!is.na(entity_clean),
         entity_clean != "",
         entity_clean != "NA") %>%
  distinct(entity_clean, eid, .keep_all = TRUE)

gbif_results <- f_get_gbif_backbone_ids_from_df(
  df = data_pubs_long,
  species_col = "entity_clean",
  kingdom = settings$gbif_kingdom,
  use_progress = FALSE,
  save = FALSE
)

# OPTIONAL FILTER
gbif_results <- gbif_results %>%
  filter(kingdom_gbif == settings$gbif_kingdom)

# JOIN BACK
gbif_results_ai <- gbif_results %>%
  left_join(data_pubs_long,
            by = c("user_supplied_name" = "entity_clean"))

# SAVE
saveRDS(
  gbif_results_ai,
  file = file.path(
    settings$out_dir,
    paste0("gbif_results_", Sys.Date(), ".rds")
  )
)
```

In these abstracts there were `r nrow(data_pubs_long %>% filter(on_topic_parsed=="yes") %>% select(eid) %>% distinct())` articles regarded as relevant. There were `r length(unique(data_pubs_long$weed_name_parsed))` unique names in the parsed name list before using the backbone taxonomy and `r length(unique(gbif_results_ai$species_gbif))` accepted species names after standardisation to the backbone taxonomy and restricting to species level.

## Table of articles per species

In our usual application of this we count the number of articles per species in our corpus [@tbl-unique-article-ids-per-species].

```{r}
#| label: tbl-unique-article-ids-per-species
#| tbl-cap: "Number of unique articles per species"


p_table_ai <- gbif_results_ai %>%
  filter(!is.na(species_gbif)) %>%
  group_by(species_gbif) %>%
  summarise(
    publication_count = n_distinct(eid),
    names_used = paste(sort(unique(user_supplied_name)), collapse = "; "),
    .groups = "drop"
  ) %>%
  arrange(desc(publication_count))

p_table_ai %>%
  gt() %>%
  tab_header(title = "Publication counts by accepted species (GBIF)")

saveRDS(
  p_table_ai,
  file = file.path(settings$out_dir,
                   paste0("article_counts_by_species_", Sys.Date(), ".rds"))
)


```

If you want to modify the prompt for a different set of questions you can potentially use an LLM provider to create new prompts in the correct format, paste the prompt that works and suggest new prompts. Or you can run another example I provided for prompt_2 (after changing the output directory).
